### 2025-07-01 paper/quarto_aaai/paper.tex
Improve formatting of equations

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
\usepackage{bibentry}
% END REMOVE bibentry
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
perform well on the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
(SGLD) with an uninformative prior for initialization but we depart from
their methodology. Instead we propose to leverage counterfactual
explainers to generate counterfactuals of observed training samples.
Specifically, we have:

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we work with a different distinction between CE and AE than
\citet{freiesleben2022intriguing} who considers misclassification as the
distinguishing feature of adversarial examples. One of the key
observations of this work is that we can leverage CEs during training
and get AEs essentially for free to reap the aforementioned benefits of
adversarial training.

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). Thus, we focus primarily on the plausibility
and cost of faithfully generated counterfactuals at test time. To
measure the cost, we follow the standard convention of using distances
(\(\ell_1\)-norm) between factuals and counterfactuals as a proxy. For
plausibility, we assess how similar CEs are to the observed samples in
the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
metric used by \citet{altmeyer2024faithful},

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

>>>>

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
% \usepackage{bibentry}
% END REMOVE bibentry
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
% For mathematical environments, you may reduce fontsize but not below 6.5 point.
\DeclareMathSizes{10}{9}{7}{6.5}

@@

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
perform well on the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
(SGLD) with an uninformative prior for initialization but we depart from
their methodology. Instead we propose to leverage counterfactual
explainers to generate counterfactuals of observed training samples.
Specifically, we have: \begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\citep{balashankar2023improving, luu2023counterfactual} or adversarial
examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \citet{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we work with a different distinction between CE and AE than
\citet{freiesleben2022intriguing} who considers misclassification as the
distinguishing feature of adversarial examples. One of the key
observations of this work is that we can leverage CEs during training
and get AEs essentially for free to reap the aforementioned benefits of
adversarial training.

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). Thus, we focus primarily on the plausibility
and cost of faithfully generated counterfactuals at test time. To
measure the cost, we follow the standard convention of using distances
(\(\ell_1\)-norm) between factuals and counterfactuals as a proxy. For
plausibility, we assess how similar CEs are to the observed samples in
the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
metric used by \citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

### 2025-07-01 paper/sections/experiments.qmd
Improve formatting of equations

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,

$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,

$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

>>>>

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
